Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-09-01 jap.manuscript.RMD
updates

##### RESEARCH ARTICLE

RUNNING HEAD: Glucose ingestion and resistance training in young adults

@@

###### Kristian Lian^1^, Sara Christine Olsen Moen^1^, Håvard Hamarsland^1^, Daniel Hammarström^1^, Stian Ellefsen^1^

<div custom-style="correspondence_style">
^1^Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inladn Norway University of Applied Sciences, Lillehammer, Norway
</div>

@@

Muscle hypertrophy, induced by resistance training (RT), is determined by ribosomal biogenesis. Synthesis of ribosomes may augment the transcription of ribosomal RNA, as evident from in vitro studies<!-- what does this actually mean?-->. However, its effect on RT-induced adaptations remains unexplored. Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 23.7 (1.8)/24.6 (4.8)), participated in a within-participant trial with unilateral resistance training with glucose- and placebo supplementation on alternate days over six sessions per condition. Participants remained fasted overnight, receiving glucose/placebo and protein supplementation only before and after RT sessions (3 sets of 10 repetitions maximum unilateral leg press and knee extension). Micro biopsies were sampled from m. vastus lateralis before and after the intervention, and maximal unilateral isometric and isokinetic knee extension torque were measured before, during (days 4, 5, 8, 9) and after the intervention. Between conditions, there were no significant differences in increases of total RNA and rRNA (mean difference `r total_rna_stats`, 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`), or levels of c-Myc, UBF and RPS6 <!-- abbreviations ok here?--> (*p* = `r min(p_west)`-`r max(p_west)`). There was a linear relationship between UBF and total RNA, where an increase of 1 SD unit of UBF equated to a $\sim$ `r corr_incr`% increase in total RNA (*p* = `r corrp`). In conclusion, glucose supplementation before and after RT does not augment RT-induced ribosome biogenesis after two weeks of heavy RT in moderately trained young adults.

@@

## NEW AND NOTEWORTHY
The effect of glucose ingestion on resistance training-induced ribosome biogenesis is quite unexplored in human skeletal muscle, and our results show no beneficial acute effect following six bouts of exercise. The linear relationship observed between UBF and total RNA do solidify the importance of UBF in human skeletal muscle ribosome biogenesis.

@@

**Keywords:** Glucose; hypertrophy; resistance training; ribosome; skeletal muscle

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition affecting the internal physiological milieu and subsequently affecting the ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Rewrite this somewhat. Focus more on what can be done to alter the responsiveness to RT, rather than genetics. I.e. make the intervention seem more sensible --> Although RT program modifications such as altered exercise volume or intensity may induce more favourable adaptations in some individuals, these benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017]. For instance, although increases in training volume generally induce greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seem necessary to optimise individual responses.

Nutritional supplements such as protein and creatine effectively optimise RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via ATP synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target S6K1 serve as a mediator for insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to test the hypothesis that glucose supplementation before and after RT increases markers of ribosomal abundance following five RT sessions. Secondly, we aimed to describe the association between UBF and total RNA abundance in human skeletal muscle. There have been multiple studies in the last years suggesting translational capacity to be as important, if not more important than,  translational efficiency with regard to long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

@@

Sixteen healthy male and female participants (20-33 years, Table 1) were recruited to the study through social media advertisement and word of mouth. The eligibility criteria were non-smokers and moderately trained (i.e. 2-8 RT sessions per 14 days for the last six months). Exclusion criteria were previous injury leading to impaired strength, inability to perform resistance exercise training, symptoms, and a medical record of metabolic disorders including hyperglycemia. Of the sixteen participants that commenced the intervention, three participants dropped out. One due to sickness and inability to resume, two participants experienced muscular discomfort related to heavy resistance training. Lean mass (Table 1) was measured using Dual-Energy X-ray Absorptiometry (Prodigy Advance PA+302047, Lunar, San Francisco, CA, USA) on Day -1, the last day preceding the RT intervention.

@@

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and -c-peptide levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources. If participants did conduct exercise training outside of the prescribed protocol, they were asked to ensure equal loading on pairwise consecutive days <!-- we do not have any data on whether this was done ore not, do we leave it out?-->.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours after RT, II) from 2.5 hours after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose/placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of PRO supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously.  During sessions, participants were free to ingest water ad libitum.

>>>>

@@

###### Kristian Lian^1^, Daniel Hammarström^1^, Håvard Hamarsland^1^, Sara Christine Olsen Moen^1^, Stian Ellefsen^1^

<div custom-style="correspondence_style">
^1^Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, Inland Norway University of Applied Sciences, Lillehammer, Norway
</div>

@@

*Purpose:* Muscle hypertrophy, induced by resistance training, is determined by ribosomal biogenesis. High-glucose treatment may augment the transcription of ribosomal RNA, as evident from in vitro studies. However, its effect on resistance training-induced adaptations remains unexplored.

*Methods:* Sixteen healthy, moderately trained individuals (male/female, n = 9/7; age, 23.7 (1.8)/24.6 (4.8)), participated in a within-participant trial with unilateral resistance training with glucose- and placebo supplementation on alternate days over six sessions per condition. Participants remained fasted overnight, receiving glucose/placebo and protein supplementation only before and after RT sessions (3 sets of 10 repetitions maximum unilateral leg press and knee extension). Micro biopsies were sampled from m. vastus lateralis before and after the intervention.

*Results:* Between conditions, there were no significant differences in increases of total RNA and ribosomal RNA (mean difference `r total_rna_stats`, 47S, `r stat_rrna$trg_47s`; 18S, `r stat_rrna$trg_18s`; 28S, `r stat_rrna$trg_28s`; 5.8S, `r stat_rrna$trg_5.8s`; 5S, `r stat_rrna$trg_5s`), or levels of c-Myc, upstream binding factor and ribosomal protein S6 (*p* = `r min(p_west)`-`r max(p_west)`). There was a linear relationship between upstream binding factor and total RNA, where an increase of 1 standard deviance unit of upstream binding factor equated to a $\sim$ `r corr_incr`% increase in total RNA (*p* = `r corrp`).

*Conclusion:* Glucose supplementation before and after RT does not augment RT-induced ribosome biogenesis after two weeks of heavy RT in moderately trained young adults.

@@

*Keywords:* Glucose; hypertrophy; resistance training; ribosome; skeletal muscle

@@

# Statements and Declarations
????

# Abbreviations
Make a table for this

@@

Responses to systematic resistance training (RT) vary widely between individuals, with as much as 10-15% showing impaired skeletal muscle growth in response to standardized training interventions [@alvarez_interindividual_2018; @mann_high_2014; @thalacker-mercer_cluster_2013]. While there certainly may be differences in the genetic predisposition in terms of readiness to adapt to exercise [@thalacker-mercer_cluster_2013], there are also recent studies showing that epigenetic factors such as ribosomal DNA (rDNA) availability may also impact the ability to mount an anabolic response to RT [@figueiredo_genetic_2021]<!--check this source -->. Indeed, the ability to synthesise ribosomes following a period of RT is connected to the magnitude of RT-induced responses in an individual [@hammarstrom_benefits_2020]. For instance, increasing training volume generally induce greater ribosome biogenesis and thus, greater benefits of RT [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_strength_2017]. However, as evident from Hammarström et al. (2020), not all participants experienced increased muscle mass and -strength despite increasing training volume. Therefore, other means than modification of RT variables alone seem necessary to optimise individual responses to RT.

Nutritional supplements such as protein and creatine effectively optimise RT adaptations [@cermak_protein_2012;@lanhers_creatine_2015;@lanhers_creatine_2017;@morton_systematic_2018]. However, it remains unknown if other nutritional adjuvants such as glucose can increase the efficacy of RT. This is surprising since glucose is the preferred energy substrate of the contracting skeletal muscle during strenuous exercise and a major energy supplier to cells via adenosine triphosphate (ATP) synthesis [@mul_exercise_2015;@sebata_control_2018]. Furthermore, energy availability is a decisive factor in the *de novo* synthesis of ribosomes [@kusnadi_regulation_2015;@moss_housekeeper_2007;@sebata_control_2018] which in turn determines muscle growth by increasing the muscle's translational capacity [@figueiredo_regulation_2019;@hammarstrom_benefits_2020;@sebata_control_2018;@stec_ribosome_2016;@von_walden_ribosome_2019]. Therefore, investigating the effects of combined RT and glucose ingestion may provide valuable insight into the potential additive effect of glucose and RT on ribosome biogenesis.

Transcription of ribosomal RNA (rRNA) by RNA Polymerase I is considered the rate-limiting step in synthesising new ribosomes [@moss_promotion_1995].  Multiple signalling pathways converge to regulate rRNA transcription, including the mammalian target of rapamycin (mTORC1) signal-transduction pathway and c-Myc [@kusnadi_regulation_2015;@von_walden_ribosome_2019;@sebata_control_2018; @west_acute_2016; @mori_c-myc_2021]. mTOR receives input from growth factors, hormones, mechanical loading, and nutrients to balance protein synthesis through multiple mechanisms based on cellular energy levels [@hoppe_amp-activated_2009]. Importantly, mTOR and its downstream target ribosomal protein S6 kinase beta-1 (S6K1) serve as a mediator for insulin signalling in skeletal muscle [@hillier_physiological_2000]. The mTORC1 pathway contributes to ribosome biogenesis through the regulation of the translation of ribosomal proteins and to forming of the preinitiation complex (PIC) that marks the initiation of rRNA transcription [@figueiredo_regulation_2019;@von_walden_ribosome_2019]. Parallel to mTORC1 [@mori_c-myc_2021; @west_acute_2016], c-Myc increases ribosomal biogenesis directly and through transcriptional control of the upstream binding factor (UBF) [@sanij_ubf_2008; @poortinga_c-myc_2011]. Additionally, UBF phosphorylation, required for interaction with the rDNA promoter, is increased by high glucose in a mTORC1 dependent manner (rapamycin sensitive) in kidney glomerular epithelial cells[@mariappan_ribosomal_2011]. Independently from UBF and mTORC1, high glucose was shown to lead to chromatin remodelling, which promoted rRNA transcription in cell cultures[@zhai_human_2012]. Together these observations indicate a potential role of glucose in positively affecting muscle rRNA synthesis.

The purpose of this investigation was to test the hypothesis that glucose supplementation before and after RT increases markers of ribosomal abundance following five RT sessions. Secondly, we aimed to describe the association between UBF and total RNA abundance in human skeletal muscle. There have been multiple studies recently suggesting translational capacity to be as important, if not more important than, translational efficiency with regard to long-term skeletal muscle adaptations to RT [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022;@figueiredo_revisiting_2019]. Previously, *in vitro* studies have shown UBF and c-Myc to be important factors in the PIC, c-Myc as a general transcription factor and UBF as a specific transcription factor for rDNA transcription initiation [@mariappan_ribosomal_2011;@von_walden_ribosome_2019]. Coupled with the recent observations by Hammarström et al. [@hammarstrom_benefits_2020;@hammarstrom_ribosome_2022], understanding underlying factors regulating and affecting ribosome biogenesis seems key to furthering our understanding of the optimisation of RT to individual phenotypes.

@@

Sixteen healthy male and female participants (20-33 years, Table 1) were recruited to the study through social media advertisement and word of mouth. The eligibility criteria were non-smokers and moderately trained (i.e. 2-8 RT sessions per 14 days for the last six months). Exclusion criteria were previous injury leading to impaired strength, inability to perform resistance exercise training, symptoms, and a medical record of metabolic disorders including hyperglycemia. Of the sixteen participants that commenced the intervention, three participants dropped out. One due to sickness and inability to resume, two participants experienced muscular discomfort related to heavy resistance training. Lean mass (Table 1) was measured using Dual-Energy X-ray Absorptiometry (DXA, Prodigy Advance PA+302047, Lunar, San Francisco, CA, USA) on Day -1, the last day preceding the RT intervention.

@@

All participants completed six RT sessions with glucose and six with placebo, allowing a within-subjects analysis of the effects of glucose ingestion before and after RT. Data from the first five RT sessions was used to investigate main outcome measures (total RNA, rRNA and protein) and leg muscle strength, whereas data from the sixth RT session was used to explore secondary outcomes (muscular recovery, plasma glucose and -c-peptide levels). Participants were asked to avoid  resistance- or high-intensity training of the legs from Day -7 (Figure 1A) and onwards, until completion of the intervention and post-testing, to ensure the reliability of pre-intervention strength data and minimal interference from external exercise sources.

@@

The dietary intervention spanned the whole day, divided into three periods: I) From awakening until 2.5 hours (hrs) after RT, II) from 2.5hrs after RT until 2200hrs, and III) from 2200hrs until awakening. During period I, participants ingested protein and glucose **or** protein and placebo only. Glucose/placebo was ingested three times in period I: 30 minutes before RT (0830hrs, 30g vs 0g glucose), immediately before RT (0900hrs, 30g vs. 0g glucose), and immediately after RT (~0930hrs, 30g vs. 0g glucose). Whey Protein Isolate (Proteinfabrikken, Stokke, Norway) was ingested 2hrs before RT (0700hrs) and immediately after RT (~0930hrs), in boluses of 25 g mixed with 150 ml water. In the afternoon (1800hrs-1900hrs, period II) participants ingested glucose or placebo (3x30g vs. 3x0g glucose) opposite to the supplement they received during RT, to ensure a balanced daily intake of glucose. Apart from this, participants ingested a self-chosen diet during period II, registered in MyFitnessPal or similar applications. The self-chosen diet was repeated on pairwise consecutive days, to ensure similar premises for resistance training responses between conditions. During period III (2200-0700hrs), participants remained in an overnight fasted state. The daily onset of the dietary intervention (i.e., first ingestion of protein supplement) varied between participants, from 0600hrs to 0900hrs to allow multiple participants to complete the protocol simultaneously. During sessions, participants were free to ingest water ad libitum.
